---
title: "Gov 50 Final Project"
author: "Bridget Galloway"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---



## Introduction and Research Question

  I am interested in exploring data related to climate change. There are so many questions to still be answered regarding the impacts of climate change; I think some that might be especially interesting to look into would be the migration patterns of areas that are most at risk for climate related disasters, or potentially the instances of health related impacts of climate change, both negative and positive. Some research has showed that the rate of respiratory illness will go down as we shift towards cleaner energy, but there still might be other negative impacts on health from climate change. 
  The following project will look into how one's environment impacts respiratory illness. The WHO produced a report that states that working to reduce climate change will simultaneously lower the risk of respiratory illnesses (https://www.who.int/news-room/fact-sheets/detail/climate-change-and-health). In this analysis, I will examine the relationship between the amount of air pollutants and the number of deaths caused by respiratory illness. I hypothesize that areas with higher amounts of air pollution are more at risk for respiratory illness. I secured data from the EPA which includes the annual and seasonal concentration reports, along with the rate of deaths caused by respiratory illness by state. While there are some variables I cannot control for, I am looking to see if there is a relationship between higher amounts of pollution in a state, and higher death rates. Because this a naturalistic observational study, I cannot control for other variables like underlying health conditions, smoking, and other things, I cannot determine a causal relationship, but I hope to find a correlation between the two. 

## Data and Visualizations
  I found data that I think can successfully prove a correlation between the environment and respiratory health. From the EPA, I gathered a data set with the AQI scores of different states across the country, and I joined that data set with another that includes the number of respiratory deaths per state from the last several years. If my hypothesis is correct, I expect to see states with worse AQI scores to experience higher rates of respiratory death. Above are my first visualizations of the data. The easiest to one to digest, based on the makeup of my data, I believe, is the tibble which includes the joined data, as it highlights both the AQI and the death rate per state, which is harder to see on the graphs. 
  The research that I will conduct with this data will be a naturalistic observational study because I am observing a naturally assigned treatment (air quality), and determining its impact on respiratory illness. Because the sample of this study is so large - the entire country is part of the sample - the external validity is quite strong. However, as I alluded to previously, this study will not be able to determine an entirely causal relationship between air quality and respiratory illness because I could not find data to properly synthesize all other confounding variables like smoking, family history with respiratory distress, and other environmental factors that may impact the number of deaths caused by respiratory illness. 
  My dependent variable is exhibited in the graph above, labeled as the resp_death_plot2020. This plot shows the respiratory death rate by state. In this research study, I am looking to show a correlation between how a state's AQI score might impact the respiratory death rate by state. In this plot, it is evident that the death rates of the states are highly varied. 

```{r}
library(tidyverse)
library(readr)
aqi_2020 <- read_csv("annual_aqi_by_county_2020.csv")
aqi_2019 <- read_csv("annual_aqi_by_county_2019.csv")
aqi_2018 <- read_csv("annual_aqi_by_county_2018.csv")
resp_death_2020 <- read_csv("respiratory_deaths_2020.csv")
```

```{r}
##BASIC PLOTS
aqi_state_mean2020 <- aqi_2020 |>
  group_by(State) |>
  summarize(state_mean = mean(`Good Days`)) 
aqi_state_mean_plot2020 <- aqi_state_mean2020 |>
  ggplot(aes(x = State, y = state_mean)) +
  geom_point()
aqi_state_mean2020
aqi_state_mean_plot2020

resp_death_plot2020 <- resp_death_2020 |>
  ggplot(aes(x = STATE, y = DEATHS)) +
  geom_point()
resp_death_plot2020
```

```{r}
##Joining the Data
library(dplyr)
aqi <- rbind(aqi_2020, aqi_2019, aqi_2018)
aqi <- aqi |>
  group_by(State, Year) |>
  summarize(state_mean = mean(`Good Days`))
respdeath2020_2 <- resp_death_2020 |>
  select(YEAR, STATE, DEATHS) |> 
  rename(Year = YEAR)
aqi <- aqi |>
  mutate(State_abb = setNames(state.abb, state.name)[State])
respdeath2020_2 |>
  colnames()
dat <- aqi |>
  left_join(respdeath2020_2, by = c("State_abb" = "STATE", "Year"))
```

## My Data, And Some Visualizations
  I found data that I think can successfully prove a correlation between the environment and respiratory health. From the EPA, I gathered a data set with the AQI scores of different states across the country, and I joined that data set with another that includes the number of respiratory deaths per state from the last several years. If my hypothesis is correct, I expect to see states with worse AQI scores to experience higher rates of respiratory death. Above are my first visualizations of the data. The easiest to one to digest, based on the makeup of my data, I believe, is the tibble which includes the joined data, as it highlights both the AQI and the death rate per state, which is harder to see on the graphs. 
  The research that I will conduct with this data will be a naturalistic observational study because I am observing a naturally assigned treatment (air quality), and determining its impact on respiratory illness. Because the sample of this study is so large - the entire country is part of the sample - the external validity is quite strong. However, as I alluded to previously, this study will not be able to determine an entirely causal relationship between air quality and respiratory illness because I could not find data to properly synthesize all other confounding variables like smoking, family history with respiratory distress, and other environmental factors that may impact the number of deaths caused by respiratory illness. 
  My dependent variable is exhibited in the graph above, labeled as the resp_death_plot2020. This plot shows the respiratory death rate by state. In this research study, I am looking to show a correlation between how a state's AQI score might impact the respiratory death rate by state. In this plot, it is evident that the death rates of the states are highly varied. 
  The independent variable that I will use for my regression is the mean of the "Good Days" variable from the AQI data set. Basically, this variable represents the average number of days a state has with a "good" AQI score. In other words, this variable counts how many days out of the year a state experiences clean air. I have named this variable state_mean. 
    
  
  
## Regression
I will now run a regression with state mean aqi as my first independent variable, and year as my second independent variable, with the number of deaths in a state per year as the outcome variable. 

```{r}
library(broom)
reg1 <- lm(DEATHS ~ state_mean, data = dat)
tidy(reg1, conf.int = TRUE, conf.level = 0.95)
mult_reg1 <- lm(DEATHS ~ state_mean  + Year, data = dat)
tidy(mult_reg1, conf.int = TRUE, conf.level = 0.95)
mult_reg2 <- lm(DEATHS ~ state_mean  + factor(Year, levels = c(2018, 2019, 2020)), data = dat)
tidy(mult_reg2, conf.int = TRUE, conf.level = 0.95)

```

```{r}
library(modelsummary)
var_labels <- c(
  "(Intercept)" = "Intercept",
  "state_mean" = "State AQI Mean",
  "Year" = "Year",
  "factor(Year, levels = c(2018, 2019, 2020))2019" = "Year 2018 v. 2020",
  "factor(Year, levels = c(2018, 2019, 2020))2020" = "Year 2018 v. 2019"
)
modelsummary::modelsummary(list(mult_reg1, reg1, mult_reg2), 
                           coef_map = var_labels,
                           gof_map = c("nobs", "R2", "R2 adj."))

```

## Interpreting My Regressions
  The first regression that I ran, labeled as reg1, used state_mean as the independent variable to predict the outcome variable, DEATHS. In other words, I was trying to find if state aqi means have a significant impact on the number of respiratory deaths in a state. The "estimate" that was produced was -9.24, which essentially means that for every 1 unit increase in the aqi variable (meaning an additional day with a good AQI score), that state is expected to have 9.24 fewer respiratory-illness related deaths.  I found, however, that the estimate of regression coefficient for state aqi has a p value of 0.184, which is outside the significance level of 0.05. I therefore must reject the null hypothesis for this regression, meaning that state aqi does not have a statistically significant impact on state deaths. 

  The second regression that I ran, which was labeled as mult_reg1, used both state_mean as well as the years variable as explanatory variables, to again predict the outcome variable, DEATHS. In this regression, I was trying to control for time since this is a potential confounding variable, to see if we can still observe statistically significant relationship between our main independent and outcome variables in case we control for the time trend. I once again found that this regression has a p value greater than 0.05, indicating that both state_mean and years do not have a statistically significant impact on the number of respiratory deaths in a state. This regression, had it been successful, would have been more likely to prove a more robust statistical test between the variables because it has a control for an additional variable, whereas my first regression did not. 

  The third and final regression that I ran, labeled mult_reg2, was similar to mult_reg1 in that it used state_mean and the years variable as explanatory variables in order to predict the outcome variable of DEATHS, but in this regression, I chose to utilize the factor function to make the years variable into a factor variable, to allow for a different relationship over time. I hoped that this would control for the potential variance between years in a state, to see if it may have a significant impact in the number of respiratory deaths in a state. Like the two regressions before this, this regression still had a p value outside of 0.05, which means that once again, the two explanatory variables do not have any statistical significance on the number of respiratory deaths. I cannot, therefore, reject the null hypothesis, meaning that I cannot accept the hypothesis that state AQI has any statistically significant effect on the number of respiratory deaths in that state. 


## Conclusions

  In conclusion, none of my three regressions were conclusive enough to reject the null hypothesis of the study; in other words, my regressions could not prove that there is a correlation between air quality, and more specifically the number of days in a year with a good AQI, and the number of respiratory related deaths per state. I think the limitations of this study were that it did not have a high enough internal validity, and therefore we could not determine causal effects of the number of days with good AQI. I think that in order to improve the study, I would like to study the AQI and death rate by county rather than state. I think that these smaller environment sizes would be more relevant and have a clearer impact because it would be more focused on the local effects of AQI on respiratory illness. Another reason that the study might lack validity is because the causes of respiratory illness are difficult to discern; in other words, they could be caused by family medical history, smoking, local environment, and other factors. In order to improve this, I would use a smaller amount of counties and try to evaluate the medical pre-existing conditions of the respiratory deaths in the county using death records. 
  
## Works Cited
https://aqs.epa.gov/aqsweb/airdata/download_files.html#Annual 
https://www.who.int/news-room/fact-sheets/detail/climate-change-and-health
https://www.cdc.gov/nchs/pressroom/sosmap/lung_disease_mortality/lung_disease.htm
